Program Transformations for miniKanren

### 2020-05-23 papers/miniKanren/text/nonconj.tex
Unfolding

Unfolding is a process of substitution of some relation call by its body with simultaneous computation of unifications.
To unfold a relation call we do the following steps.
\db{TODO: the following has alredy mentioned above.}
First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
All fresh variables get instantiated.
The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
All unifications are computed.
Those disjuncts in which unifications fails are removed.
Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.

The most important question is when to unfold.
\db{
  {\color{blue}PROPOSED EXTENSION:}, and when to stop unfolding even if the relation call seems to be good for specialization.
  Usually, unfolding in functional programming languages specialization, as well as inlining in imperative one, is considered to be safe from a residual program efficiency point of view.
  It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if specializer is considered as a standalone tool~\cite{jonesbook}.
  Unfortunately, this is not the case for the specialization of relational programming language.
  Unlike functional and imperative, in relational programming language unfolding may easily affect the target program efficiency.
  It is crucial not to unfold too much since, for example, extra unifications or even repeated computations generated by optimized relation calls and further propagated to neighbouring conjunction may appear.
}
\db{{\color{blue}Unfortunately $\to$ }That is}, too much unfolding is sometimes is even worse than not enough unfolding.
There is a fine edge between those.
This problem is mentioned in the (CONTROL PAPER).
We believe that the following heuristic provides a reasonable control.

>>>>

Unfolding in our case is done by substitution of some relation call by its body with simultaneous normalization and computation of unifications.
% To unfold a relation call we do the following steps.
% \db{TODO: the following has alredy mentioned above.}
% First, the formal arguments of a relation are substituted for the actual arguments of the call in the body.
% All fresh variables get instantiated.
% The body is transformed into a canonical form (disjunction of conjunctions of either calls or unifications).
% All unifications are computed.
% Those disjuncts in which unifications fails are removed.
% Other disjuncts take form of a conjunction of relation calls accompanied with a substitution.
The unfolding itself is straightforward however it is not always clear what to unfold and when to \emph{stop} unfolding.
Unfolding in functional programming languages specialization, as well as inlining in imperative one, is usually considered to be safe from a residual program efficiency point of view.
It may only lead to code explosion or code duplication which is mostly left to a target program compiler optimization or even is out of consideration at all if a specializer is considered as a standalone tool~\cite{jonesbook}.

Unfortunately, this is not the case for the specialization of relational programming language.
Unlike functional and imperative, in logic and relational programming language unfolding may easily affect the target program efficiency~\cite{leuschel2002logic}.
Unfolding too much may create extra unifications, which is by itself a costly operation, or even introduce duplicated computations by propagating the unfolding results onto neighboring conjuncts.

There is a fine edge between too much unfolding and not enough unfolding.
The former may be even worse than the latter.
We believe that the following heuristic provides a reasonable approach to controling unfolding.
